The Rockerverse: Images, Packages, and Applications for Containerization with R

### 2020-02-18 manuscript.Rmd
Update dockr paragraph, authors in alphabetical order

title: "The Rockerverse: Packages and Applications for Containerization with R"

@@

  The Rocker Project provides widely-used Docker images for R across different application scenarios.

@@

@@

```{r setup, include=FALSE}
Sys.setlocale('LC_ALL','C')
```

@@

The R community keeps growing.
This can be seen in the number of new packages on CRAN, which keeps on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies taking up, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers indirectly or only for specific features, up to complex use cases leveraging containers.
We conclude with a reflection on the landscape of packages and applications and point out future directions of development.

@@

For statistical computing and analysis centered around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for better maintainability (i.e., smaller `Dockerfile`s), composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

# Interfaces for Docker in R
\label{interfaces}

@@

# Use cases and applications
\label{applications}

## Image stacks for communities of practice

>>>>

title: "The Rockerverse: Images, Packages, and Applications for Containerization with R"

@@

  The Rocker&nbsp;Project provides widely-used Docker images for R across different application scenarios.

@@

  We also look beyond Rocker to other projects connecting containerisation with R, namely alternative suites of images.

@@

```{r setup, include=FALSE}
library(tidyverse)
```

@@

The R community keeps growing.
This can be seen in the number of new packages on CRAN, which keeps on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies taking up, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the _Claerbout/Donoho/Peng_ terminology \citep{barba_terminologies_2018}.} to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by descriptions of projects developing images with R installations.
Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers indirectly or only for specific features, up to complex use cases leveraging containers.
We conclude with a reflection on the landscape of packages and applications and point out future directions of development.

@@

For statistical computing and analysis centered around R, the Rocker&nbsp;Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for better maintainability (i.e., smaller `Dockerfile`s), composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

# Container images
\label{images}

## Images for alternative R engines
\label{alternatives}

As outlined above, R is a widely-used language with a large community.
The large number of extension packages provides access to an unrivaled variety of established and upcoming features.
Nevertheless, special use cases and experimental projects in academia and industry exist to test new approaches, or provide features different to what the R implementation maintained by the [R Core Team](https://www.r-project.org/contributors.html) and available via [CRAN](https://cran.r-project.org/) provides.
While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to others.
All these alternative implementations of R are published under GPL licenses.
Since all implementations of R claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.

**Microsoft R Open** (MRO) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and its packages and it "[..] includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository, which is enabled by default, and some bundled packages such as \CRANpkg{checkpoint} and specific R packages maintained by Microsoft\footnote{See "Bundled Packages" on \href{https://mran.microsoft.com/rro/installed}{https://mran.microsoft.com/rro/installed}.}.
Originally the optional integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread} was a distinguishing feature before it was made available publicly, and can be used in base R, too \citep{eddelbuettel_thinking_2018}.

MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g., the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

**Renjin** is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability and integration into enterprise platforms.
R extension packages need to be specially compiled and are distributed via the Java package manager [Apache Maven](https://en.wikipedia.org/wiki/Apache_Maven), cf. [http://packages.renjin.org/packages](http://packages.renjin.org/packages) for available packages.
Packages are loaded on demand, i.e., at the first call to `library()`.
Not all R packages, especially one linking to binary libraries, are available, e.g., `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
These images expose the command line interface of Renjin in a similar fashion as Rocker images and allow an easy evaluation of Renjin's suitability, but are not intended for production use.

**pqR** ([http://www.pqr-project.org/](http://www.pqr-project.org/)) is a "a pretty quick version of R".
_pqR_ attempts to improve R on some opinionated issues in the R language and is the basis for experimental features, e.g., automatic diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` image was independently created, see Docker Hub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and GitHub at [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker).

**FastR** ([https://github.com/oracle/fastr](https://github.com/oracle/fastr)) is "A high-performance implementation of the R programming language, built on GraalVM" \citep{oracle_labs_oraclefastr_2020}.
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \citep{oracle_labs_oraclefastr_2020}.
There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g., [`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).

## Bioconductor
\label{bioc}

_Bioconductor_ ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
The project consists of 1823 R software packages as of October 30th 2019, as well as packages containing annotation or experiment data.
_Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R, and Docker images are provided for current and past versions of _Bioconductor_ for convenience and reproducibility.
All images, included  are described on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub, and distributed through Docker&nbsp;Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
_Bioconductor_'s 'base' Docker images are built on top of the `rocker/rstudio` image.
_Bioconductor_ installs packages based on the R version in combination with the Bioconductor version, and therefore uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
Past and current combinations of R and _Bioconductor_ will therefore be accessible via a specific image tags.
The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages.
A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
A recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to extent.
As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages and saves users and package developers from managing these themselves.
Furthermore the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
Users mounting a location on the host file system to this location can persist installed packages across Docker sessions or image updates.
Many R users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
The new `bioconductor_docker` image is well-suited for this pattern, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.

## Images for (historic) R versions
\label{versions}

As with any other software, each new version of R comes with its share of changes.
Some are breaking changes, some are not, but the fact is that running a piece of code in a given version might give different results from another version can be an issue when it comes to reproducibility of workflows and stability of applications.
For example, think about R random seed: starting with `R 3.6.0`, running `set.seed(2811); sample(1:1000, 2)` will not give the same result as if it was run inside an older version of R.
That might seem trivial, but all the code using the random number generator will not be exactly reproducible after this breaking change.
Therefore, controlling the version of software is most crucial for reproducible research \citep[e.g.][]{boettiger_introduction_2015}.

Containers are perfectly suited to capture a specific configuration of a computing environment to prohibit such problems.
The Rocker images provides the _versioned stack_ for different the version of R since the project inception, which use the _then-current_ stable Debian base image \citep[cf.][]{RJ-2017-065}.
Based on custom build phase hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}}, i.e., small shell scripts executed at different phases in the automated build of Rocker images, there are also semantic version tags for the most recent freezed (i.e., using MRAN) versions with only the major and minor version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42}{https://github.com/rocker-org/rocker-versioned/issues/42}.}.
So, at the time of writing this article, `rocker/r-ver:3` and `3.6` are aliases for `3.6.0`, because `3.6.1` is the latest release.
With the release of `3.6.2` and pinning of the MRAN version in `3.6.1`, the tags `3` and `3.6` will deliver the same image as `3.6.1`.
These tags allow users to update the base image to retrieve bugfixes while reducing the risk of also introducing breaking changes.

[r-online](https://github.com/ColinFay/ronline) is an app for helping users to detect breaking changes between different R versions, and for historic exploration of R.
With a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.

A container also provides an isolated sandbox environment suitable for testing and evaluation, without interfering with the "main" working environment.
This enables cross version testing and bugfixing.
To take this even further, the Rocker contributors are discussing the provision of R versions reaching back futher than the project's own inception, reaching back as far as R `2.x` and `1.x` \footnote{\href{https://github.com/rocker-org/rocker-versioned/issues/138}{https://github.com/rocker-org/rocker-versioned/issues/138}}.
The main challenges are finding a suitable base image with a matching OS version, or making the adjustments to compile R on more recent OS releases.

## Windows Images
\label{windows}

Docker containers on the Windows operating system were originally quite cumbersome to use.
They required an extra tool, [Docker Toolbox](https://docs.docker.com/toolbox/), which by now only exists as a legacy solution for older Windows systems.
Docker toolbox leverages `docker-machine` (see also Section&nbsp;\nameref{interfaces} for an R package interfacing with `docker-machine`) to handle the process of creating a local virtual host which could host Docker Engine, while exposing the regular Docker CLI.
The Docker CLI commands are forwarded to the virtual host transparently for the user.

For current Windows Server (2016 and later) and Windows Desktop ([Docker Desktop](https://www.docker.com/products/docker-desktop) requires Windows 10) versions, Docker is supported natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}} and different base images are offered by Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
The Docker CLI can be used in just the same way as on other operating systems but not every base image is supported on every Windows host\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}}.
On Docker Desktop for Windows, the user can run both Linux-based and Windows-based containers, but only one of the Docker daemons can be used at a time\footnote{See "Switch between Windows and Linux containers" on \href{https://docs.docker.com/docker-for-windows/}{https://docs.docker.com/docker-for-windows/}.}.

**rocker-win** ([https://github.com/nuest/rocker-win](https://github.com/nuest/rocker-win)) is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
These base images match different Windows versions: Windows Server 2019, Windows Server 2016, and Windows Server, version 1803.
The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker&nbsp;Hub](https://hub.docker.com/r/nuest/rocker-win).
The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
All these images can be run on Docker for Desktop on Windows 10.
New R versions are only added on demand.
The explorative rocker-win project demonstrates the variety of Windows on Docker, which is potentially confusing, e.g., because of the different base images.
The proof of concept provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \CRANpkg{plumber}, \CRANpkg{Shiny}, or R packages with system dependencies such as \CRANpkg{sf}.
The latter is quite comfortable and fast actually, compared to the default installation from source on Linux, because CRAN ships pre-compiled binaries for Windows.
However, compared to the various Linux images, the footprint of Windows images is quite large.

For Windows, just as originally for Linux, the cloud use cases drive the development and Docker and Microsoft collaborate closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
Naturally, a Windows license is required for the host machine, and the licensing impacts the potential to build images in clouds or redistribute exported images as archive files.
But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.

## Non-Rocker Linux images
\label{nondebian}

\label{rhub}
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \CRANpkg{rhub} \citep{csardi_rhub_2019}.
The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
The `Dockerfile`s and images are published on Docker&nbsp;Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\nameref{development}) not for reproducible workflows.
_The rationale_ of not re-using the Rocker image stack is the high number of operatings systems and configurations, which can be more uniformly maintained in a seperate suite of images.
Also, a common structure across the respective image stacks of the operating systems, e.g., for `debian`, adding GCC in `debian-gcc` and then R's development version in `debian-gcc-devel`.
The platform also has a specific user configuration differing from the setup covering typical applications of Rocker images, and the images include considerably more software, such as a fairly complete LaTeX installation, to provide an experience closer to CRAN.

[**RStudio**](https://rstudio.com/) also maintains a set of [Docker images](https://github.com/rstudio/r-docker) for multiple operating systems including SUSE, Ubuntu, CentOS, and Debian.
The image stack includes an [opinionated R installation](https://github.com/rstudio/r-builds) and ensures the R installation and profile is consistent across the different Linux distributions.
The base image also installs R in a versioned directory and uses a minimal set of build and runtime dependencies.
Furthermore these images are responsible for creating pre-compiled binary R packages for Linux, which supplement the binaries built by CRAN.
These pre-compiled packages dramatically decrease the installation time of R packages on Linux and subsequently can decrease the build time of Docker images \citep{lopp_package_2019}.

There are also independent projects installing R or RStudio on [**Alpine Linux**](https://www.alpinelinux.org/), but not beyond the proof of concept stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
In general, Alpine-based Docker images are often chosen for minimalistic and thereby small and secure containers, e.g., in hardware with limited storage.
R-hub's proof of concept [`r-minimal`](https://github.com/r-hub/r-minimal) takes this to the extreme with compressed image size of 20MB and bespoke install scripts, e.g., to cleverly remove compilers after package installation.
For R though, this advantage quickly deteriorates if more features are needed, e.g., if adding compilers, packages, or especially if installing powerful data science and communication tools, like RStudio or LaTeX.
Furthermore the distribution uses the [`musl` C library](https://www.musl-libc.org/) instead of the `glibc` used in Debian/Ubuntu, for which more tooling and experiences exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
In fact, using Debian's 'slim' variant and removing capabilities, one could probalby achieve similarly small image sizes on a more common stack.

**\pkg{altRnative}** ([https://github.com/ismailsunni/altRnative/](https://github.com/ismailsunni/altRnative/)) is an experimental R package for running the same code across multiple containerised versions of R, intended for comparison across operating systems and different implementations of R.
It comes with a collection of `Dockerfile`s and corresponding images in multiple combinations, currently including, e.g., MRO, FastR, Fedora, and TERR\footnote{See `Dockerfile`s at \href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}.}.

## Data Science images
\label{datascience}

_Data Science_ is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
The discussions shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
The practice of Data Science often involves a combination tools and software stacks and requires a cross-cutting skillset.
This complexity and an inherent concern for openness and reproducibility in the Data Science community lead to Docker being used widely.
<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
This section presents exemplary Docker images and image stacks featuring R intended for Data Science.

The [**Jupyter Docker Stacks**](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells.
R is also included in the catchall `jupyter/datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
R is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.

**Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and also configures the \CRANpkg{reticulate} package.
The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.

As a further example, [**Radiant** project](https://radiant-rstats.github.io/docs/) provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9GB.

**Gigantum** ([http://gigantum.com/](http://gigantum.com/) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects.
It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
The independent image stack can be traced back to the Gigantum environment and its features.
The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.} so that appropriate GPU drivers can be enabled automatically when supported.
Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusions}).
Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.

**Whole&nbsp;Tale** (https://wholetale.org) is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole~Tale a _tale_ is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
Using Whole&nbsp;Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
Additionally, Whole&nbsp;Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
For R users, Whole&nbsp;Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS (long-term support) required to support the Rocker Project images. 
Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
This works in part because Rocker images are based on a Debian distribution, so the instructions created by `repo2docker` for Ubuntu work because of compatible shell and package manager.

# Use cases and applications
<!-- from packages closer to Docker to second level packages to more complex use cases -->
\label{applications}

## Interfaces for Docker in R
\label{interfaces}

@@
